\documentclass{letter}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}


%\address{Department of Computer Science, Applied Mathematics and Statistics,\\ Univesitat de Girona, P4-Campus Montilivi,\\ 17071 Girona, Spain}
\begin{document}


\begin{letter}{
Scandinavian Journal of Statistics}
\opening{Dear Editor:}

Please find enclosed the manuscript: “Logratio methods in mixture models for compositional data sets”, by Marc Comas-Cufí, Josep-Antoni Martín-Fernández, Glòria Mateu-Figueras, to be submitted as a Regular Paper to the Scandinavian Journal of Statistics for consideration for publication. All co-authors have seen and agree with the contents of the manuscript and there is no financial interest to report. We certify that the submission is original work and is not under review at any other publication.

As in the case of spatial and longitudinal data, when compositional data are analysed specific methods should be used. This type of data has a particular nature that motivates important research problems for their statistical modelling. In this manuscript we show that a finite mixture of distributions defined on the real space is not appropriate for compositional data. Traditional methods we can find in the literature are revisited and their major difficulties are detailed. A new and general proposal using a mixture of distributions defined on log-ratio coordinates is presented. Finally, to compare the different methodologies, different finite mixture distributions are adjusted to a real compositional data set.

We believe that our findings could be of interest to the readers of the Scandinavian Journal of Statistics because they bring a new methodology for modelling compositional data through the use of finite mixture distributions. 

We hope that the editorial board will agree on the interest of this study.

Sincerely yours,

\bigskip

Marc Comas-Cufí on behalf of the authors.

\noindent\rule{8cm}{0.4pt}

Address of the corresponding author: \\
Marc Comas-Cufí,
Dept. Computer Science, Applied Mathematics and Statistics, University of Girona, Campus Montilivi (P4), E-17071 Girona, Spain. (mcomas@imae.udg.edu; (+34) 972 418426)

\end{letter}

\end{document}
