---
title: "Eliminating one component"
author: "mcomas"
date: "23/07/2014"
output:
  html_document:
    theme: null
---

### The effect of eliminating one component

 Compositional data is commonly presented in such a way that the sum of each observation is some constant (1, 100, etc). When classical methods are applied, due to the colinearity between components, a common practice is to eliminate one component for the rest of the analysis. In this section, we an example is shown where although one component is eliminated from the analysis, the effect of such component is still present after its elimination.

Consider the following example:

> Suppose that a compositional sample $X$ with two components, $a$ and $b$, is originated from two different sources $S_1$ and $S_2$. 

> * $S_1$ has a higher amount of component $a$ compared to component $b$
> * $S_2$ has a higher amount of component $b$ compared to component $a$

```{r, echo=F, include=FALSE}
library(ggplot2)

N1 = 100
N2 = 100
SD = 0.25
## Random data generation
S1.comp = data.frame('a' = exp(rnorm(N1, mean = 0, sd=SD)), 
                     'b' = exp(rnorm(N1, mean =-1, sd=SD)),
                     'c' = exp(rnorm(N1, mean =-2, sd=SD)))
S2.comp = data.frame('a' = exp(rnorm(N2, mean =-1, sd=SD)), 
                     'b' = exp(rnorm(N2, mean = 0, sd=SD)),
                     'c' = exp(c(rnorm(N2/2, mean = 1, sd=SD), rnorm(N2/2, mean = -1.5, sd=SD))))

```

```{r, include=F}
## Clousuring
S1 = round(S1.comp[,c('a','b')] / apply(S1.comp[,c('a','b')], 1, sum) *100, 2)
S2 = round(S2.comp[,c('a','b')] / apply(S2.comp[,c('a','b')], 1, sum) *100, 2)
## Labels
S1$src = 'S1'
S2$src = 'S2'
X = rbind(S1, S2)
```

> Here, we can see four elements coming from source $S_1$

```{r, echo=FALSE}
X[1:4,c('src', 'a', 'b')]
```

> and four elements coming from source $S_2$

```{r, echo=FALSE}
X[c(N1+1:2, N1+N2/2+1:2),c('src', 'a', 'b')]
```

> To represent the compositional sample $X$ with components $a$ and $b$, we can use a typical $\mathbb{R}^2$ scatterplot representation

```{r, echo=F, fig.width=4, fig.height=3}
ggplot(data=X, aes(x=a, y=b, col=src)) + geom_point(size=2)+theme_bw()
```

> or we can represent the amount of part $a$ in an histogram (to better identify the sources, a histogram for each source has been ploted)

```{r, echo=F, fig.width=3.5, fig.height=3}
ggplot(data=X, aes(x=a, fill=src)) + geom_histogram(binwidth=2.5) + facet_grid(src~.) + ylab(NULL)+theme(legend.position="none")+theme_bw()
```

> **Observation:** At this point, using component $a$ and $b$, it is natural to indentify two different groups. The first one with a higher amount of element $a$ compared to $b$, and a second one with a higher amount of element $b$ compared to $a$.

> Now, suppose that another component $c$ is included as follows: for observations coming from source $S_1$, the component $c$ is *lower* compared to component $a$ and $b$, for observations coming from source $S_2$, in some cases component $c$ is *higher* that component $a$ and $b$, and in other cases component $c$ is *lower* than component $a$ and $b$.

```{r, include=F}
## Clousuring
S1 = round(S1.comp[,c('a','b','c')] / apply(S1.comp[,c('a','b','c')], 1, sum) *100, 2)
S2 = round(S2.comp[,c('a','b','c')] / apply(S2.comp[,c('a','b','c')], 1, sum) *100, 2)
## Labels
S1$src = 'S1'
S2$src = 'S2'
X = rbind(S1, S2)
```

> Here we show the same observations coming from source $S_1$ than before with component $c$ included

```{r, echo=FALSE}
X[1:4,c('src', 'a', 'b', 'c')]
```

> and the four observations coming from source $S_2$ (in the first two observations component $c$ is higher than $a$ and $b$, while in the last two observations component $c$ is lower than $a$ and $b$)

```{r, echo=FALSE}
X[c(N1+1:2, N1+N2/2+1:2),c('src', 'a', 'b', 'c')]
```


```{r, include=FALSE}
library(ggtern)
```

> A representation of sample $X$ with components $a$, $b$ and $c$.

```{r, echo=F, fig.width=5, fig.height=5}
ggtern(X, aes(x=a, y=b, z=c, col=src))+geom_point()+theme_bw()
```


> **Comment:** To think in a practical example, suppose parts a and b are elements of importance in our analisis, and part c is another residual part, for example water: 

> * The sample comming from source 1 was collected in days where the amount of water was very low. 
> * The sample coming from source 2 was collected in two different days, one with a very low level of water and another with a higher level of water (maybe, it was raining). 

*****************

### What does `eliminating a component c` mean ?

To deal with compositional data, a common approach consists in eliminating a component and then, follows with classical methods using the other components. For example, in previous example, if we decide to eliminate component $c$, we end up with components $a$ and $b$

```{r, echo=FALSE}
X[c(1:4,N1+1:2, N1+N2/2+1:2),c('src', 'a', 'b')]
```

which can be represented in $\mathbb{R}^2$ as

```{r, echo=F, fig.width=4, fig.height=3}
ggplot(data=X, aes(x=a, y=b, col=src)) + geom_point(size=2)+theme_bw()
```

It is evident that the effect of component $c$ is still present in our analysis.

If we where interested in completely eliminating the effect of component $c$, we should restrict observation to sum a constant. In this way, we can obtain again the same sample we have before including component $c$.


**Final comments:** *Eliminating a component because we have colinearity problems* (elimination e1) is different from *eliminating a component because we are interested in removing the effect of that component in our analysis* (elimination e1). Elimination e1 is just an statistic/mathematic artict which is far from the decision of eliminating a component from our analysis (elimination e2).

In next section, we introduce a methodology that when we decide to eliminate a component we are dealing with elimination e2.
