---
title: "Eliminating one component"
date: "28/07/2014"
output:
  html_document:
    theme: null
---


#### Example

```{r, echo=F, include=FALSE}
set.seed(1)
library(ggtern)
N1 = 10
N2 = 10
SD = 0.25
## Random data generation
S1.comp = data.frame('a' = exp(rnorm(N1, mean = 0, sd=SD)), 
                     'b' = exp(rnorm(N1, mean =-1, sd=SD)),
                     'c' = exp(rnorm(N1, mean =-2, sd=SD)))
S2.comp = data.frame('a' = exp(rnorm(N2, mean =-1, sd=SD)), 
                     'b' = exp(rnorm(N2, mean = 0, sd=SD)),
                     'c' = exp(c(rnorm(N2/2, mean = -1.5, sd=SD), rnorm(N2/2, mean = 1, sd=SD))))
## Clousuring
S1 = round(S1.comp[,c('a','b','c')] / apply(S1.comp[,c('a','b','c')], 1, sum) *100, 2)
S2 = round(S2.comp[,c('a','b','c')] / apply(S2.comp[,c('a','b','c')], 1, sum) *100, 2)
## Labels
S1$site = 'S1'
S2$site = 'S2'
S1$condition = 'C1'
S2$condition = rep(c('C1', 'C2'), each=N2/2)
X = rbind(S1, S2)
```

Consider a compositional sample $X$ with three components, $a$, $b$ and $c$. Suppose that each observation of $X$ was collected from two different sites, $S_1$ and $S_2$. Moreover, suppose that each observation could be collected in two different conditions $C_1$ and $C_2$ which are conditioning the level of element $c$. For example, suppose we have the following sample with $20$ observations:

<table>
<tr>
<td>
```{r, echo=F, fig.width=5, fig.height=5}
(p1<-ggtern(X, aes(x=a, y=b, z=c, col=site, shape=condition))+geom_point(size=3)+theme_bw())
ggsave(p1, file='figures/example_ternary.pdf')
```
</td>
<td>
<pre>
```{r, echo=FALSE, comment="       "}
X[,c('a', 'b', 'c', 'site', 'condition')]
```
</pre>
</td>
</tr>
</table>

Note that all the observation collected in site $S_1$ were collected under condition $C_1$. In contrast, note that some observations from site $S_2$ were collected in condition $C_1$ and some in condition $C_2$.

> To think in a practical example, we can think that condition $C_1$ is a sunny day, $C_2$ is a rainy day and component $c$ is water.

Suppose now, that a researcher is interested in fitting a mixture model to the compositional sample $X$. Because for each observation $i$, $a_i+b_i+c_i = 100$, there is a perfect co-linearity between components. Therefore, the standard multivariate normal density can not be evaluated.

Because component $c$ contains the effect of condition $C_1$ and $C_2$, the researcher decides to eliminated component $c$ from the rest of the analysis. Doing so, he also solves the co-linearity problem. The researcher ends up with sample.

<table>
<tr>
<td>
```{r, echo=F, fig.width=5, fig.height=5}
(p2<-ggplot(X, aes(x=a, y=b, col=site, shape=condition))+geom_point(size=3)+theme_bw()
```
</td>
<td>
<pre>
```{r, echo=FALSE, comment="       "}
(X2<-X[,c('a', 'b', 'site', 'condition')])
```
</pre>
</td>
</tr>
</table>

Although the researcher was interested in removing the effect of condition $C_1$ and $C_2$ by eliminating component $c$, its effect is still present after eliminating it from the rest of the analysis.
